---
title: "On the Contribution of COVID-19 compared to other Communicable Diseases towards Outbreaks in Healthcare Institutions in Ontario"
author: 
  - Owais Zahid
thanks: "Code and data are available at: https://github.com/FFFiend/outbreaks_in_toronto"
date: today
date-format: long
abstract: "This paper examines the contribution of COVID-19 to the total number of outbreaks within healthcare institutions in Ontario over all months in 2023. It was found that COVID-19 was in fact the highest contributor to outbreaks for the aforementioned year. This investigation is important because it helps healthcare institutions within the province recognize that they need to enforce stricter guidelines surrounding patient hygiene."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)

### Workspace setup ###
grab_data = read_csv(file=here("outputs/data/cleaned_data.csv"))
grab_data
```

# Introduction
The introduction of vaccines and boosters is what got us out of the COVID-19 pandemic and enabled society to return back to normal. However 4 years and a couple booster shots later, it was found that COVID-19 is still at large in Ontario, attributing to the majority of outbreaks within local healthcare institutions.

Outbreak data from various healthcare institutions (including different floors/departments within these institutions) in Ontario for the year 2023 was analyzed, and a total of 18 diseases (along with some cases left as Pending) attributing to the outbreaks were found, which will be listed out in detail in the Data section. Of all 18 possible diseases, COVID-19 was found to be the highest contributor to outbreaks in 2023 by far.

In subsequent sections, the data will be examined closely, cleaned and contributions from each disease will be analyzed over all the months in 2023. There will be more focus towards COVID-19 however, as we cannot conclude simply from it being the highest contributor overall, that it had a uniformly maximum contribution over all the months in 2023.


# Data 
The outbreak data is from "Outbreaks in Toronto Healthcare Institutions" [@citedata] on the Open Data Toronto website [@citeopentoronto]. 

All data anlaysis was done using the R programming language [@citeR], on the R-Studio platform [@citeRStudio]. Furthermore, the libraries tidyverse [@citetidyverse] and here [@citehere] were also used.

The year chosen to examine is 2023, and the raw csv has the following columns.

```{r}
#| include: true
#| echo: false
#| warning: false
#| message: false
raw_csv = read_csv(file=here("inputs/data/raw_data.csv"))
cols = colnames(raw_csv)
cols
```
The focus will be on the the "Date Outbreak Began" and "Causative Agent-1" columns. The "Causative Agent-2" column will be ignored as some entries are missing values in the aforementioned column. Since the identity of each entry (i.e the institution name, address, and outbreak setting) are of no consequence in our analysis, they will also be ignored. The "_id" variable simply refers to the actual index of the entry within the dataset, and so that is not worth exploring either. We do not concern ourselves with "Date Declared Over" as the speed at which individuals recover from COVID-19 is not the focus of this paper. 

# Results
We first begin by plotting the number of cases contributed by each of the 18 unique diseases responsible for outbreaks in 2023. The "Pending" category has also been included and the following histogram is obtained. 
```{r}
#| echo: false
ggplot(grab_data, aes(x = factor(Causative.Agent.1), fill = factor(Causative.Agent.1))) +
  geom_bar() +
  labs(title = "Figure 1: Number of Cases by Causative Agent-1 in 2023",
       x = "Causative Agent-1",
       y = "Number of Cases") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        plot.title = element_text(size = 8),  # Adjust title size
        axis.title.x = element_text(size = 8),  # Adjust x-axis label size
        axis.title.y = element_text(size = 8))
```
It can be observed in Figure 1, that COVID-19 was the highest contributor to outbreaks in 2023 by far, with exactly 700 out of the total 1066 outbreaks (or 65.66%) in Ontario. Further investigation can be done to observe the spread of COVID-19 cases throughout the year as follows:
```{r}
#| include: true
#| echo: false
#| warning: false
#| message: false
covid_data <- grab_data %>%
  filter(Causative.Agent.1 == "COVID-19" & grepl("2023", Date.Outbreak.Began))

# Convert Date.Outbreak.began to Date type
covid_data$Date.Outbreak.Began <- as.Date(covid_data$Date.Outbreak.Began)

ggplot(covid_data, aes(x = format(Date.Outbreak.Began, "%m"), fill = Causative.Agent.1)) +
  geom_bar(stat = "count") +
  labs(title = "Figure 2: Number of COVID-19 Cases per Month in 2023",
       x = "Month",
       y = "Number of Cases") +
  scale_x_discrete(labels = month.abb) +
  theme(
    legend.position="none",
    plot.title=element_text(size=8),
    axis.title.x = element_text(size = 8),  # Adjust x-axis label size
    axis.title.y = element_text(size = 8)
  )
```
It can be observed in Figure 2, that most of the COVID-19 outbreaks occurred around the start of Winter towards the end of the year. The data is left skewed, and the maximum number of outbreaks (131) was recorded in December. That being said, there is a local peak of 78 cases in January. 

Although the number of COVID-19 outbreaks starts to increase from September, it may still be the highest contributor towards outbreaks in months where the total number of cases is relatively mild in comparison.

Thus, detailed breakdown of outbreaks attributed to the 19 categories of diseases (including Pending,) within healthcare institutions between the months of February and August (ordered alphabetically) can be seen below:

```{r}
#| include: true
#| echo: false
#| warning: false
#| message: false
data_filtered <- grab_data %>%
  filter(`Date.Outbreak.Began` >= "2023-02-01" & `Date.Outbreak.Began` <= "2023-08-31") %>%
  mutate(month = format(as.Date(`Date.Outbreak.Began`, format="%m/%d/%Y"), "%B"))

# Create a histogram
ggplot(data_filtered, aes(x = month, fill = `Causative.Agent.1`)) +
  geom_bar() +
  labs(title = "Figure 3: Number of Outbreak Cases per Month in 2023",
       x = "Month",
       y = "Number of Cases") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position="bottom",
        legend.text = element_text(size = 4),  # Set legend text size
        legend.title = element_text(size = 6),
        plot.title = element_text(size=8),
        axis.title.x = element_text(size=8),
        axis.title.y = element_text(size=8))
```
It can be seen in Figure 3, that even among months where COVID-19 was seemingly not spiking (i.e in the middle of the year) it still made up the majority percentage of all outbreaks within those months, i.e it was the highest contributor as shown by the plot above.

# Conclusion
We can thus conclude that healthcare institutions in Ontario have their work cut out for themselves if they wish to reduce the number of outbreaks that occur moving forward. This may be accomplished by adding more posters on hygiene and masking etiquette in and around the institutions, adding safety screen shields at any and all help desks, and enforcing both medical staff and patients to wear face shields, or face masks at all times. 


\newpage


# References